Contracts for unions and intersections are hard for Nickel

### 2021-06-07 paper.tex
Last bit of reworking section 4

A natural - but naive - implementation of intersection contracts could be the
following: to apply a contract \nickel{A @& B}, apply both contracts \nickel{A}
and \nickel{B} sequentially, resulting in the naive decomposition rule of
Figure~\ref{fig:naiv-impl}.

\begin{figure}[h]
\label{fig:naive-impl}
\begin{lstlisting}[language=nickel,frame=none,numbers=none,title={Naive
decomposition}]
M | A @& B (*@ $\simeq$ @*) (M | A) | B
\end{lstlisting}
\begin{lstlisting}[language=nickel,frame=none,numbers=none,title={Exchange law}]
(A -> B) @& (C -> D) (*@ $\simeq$ @*) (A @& C) -> (B @& D)
\end{lstlisting}
\begin{lstlisting}[language=nickel,title={Overloaded identity}]
let g | Num -> Num @& Str -> Str
      = fun x => x in
g 1
\end{lstlisting}
\caption{Naive implementation of intersection}
\end{figure}

This intuition works for simple contracts: checking that \nickel{x | Natural @&
Odd} amounts to check that \nickel{x | Natural} and \nickel{x | Odd}.
Unfortunately, this doesn't scale to higher-order contracts.  The overloaded
identity example of Figure~\ref{fig:naive-impl} illustrates the use of an
intersection to model a simple overloading of the identity function. If we were
to apply the naive decomposition, the argument \nickel{1} would fail the
\nickel{Str -> Str} contract and abort the execution. Perhaps the exchange rule
given in Figure~\ref{fig:naiv-impl}, which is a direct consequence of the naive
decomposition, illustrates the issue better. It is clear that this exchange law
isn't the right semantics for overloading. With this law, the contract for
overloaded identity of Figure~\ref{fig:naiv-impl} would always fail because
no argument can satisfy \nickel{Num @& Str}.

@@

@@

But which semantics? We deemed the exchange law of Figure~\ref{fig:naive-impl}
as clearly wrong with respect to overloading. There is nothing wrong about it
per se : we could chose a semantics for intersections that validate this rule.
However, the resulting system wouldn't capture overloading, and is arguably not
very interesting nor useful. In~\ref{KeilThiemannUnionIntersection}, the authors
argue for basing the semantics of union and intersection contracts on union and
intersection type systems, where they have a simple and natural definition from
which the other properties follow. This approach makes sense, and has been
continued in future works, and in particular in~\ref{RootCauseOfBlame}.

But even this semantics is no panacea. Figure~\ref{fig:intersection-distrbution}
shows an example that intuitively looks like it should succeeed, returning
\nickel{(1, 1)}. However, under the semantics used both
in~\ref{KeilThiemannUnionIntersection} and~\ref{RootCauseOfBlame}, this program
fails with a contract violation.

\begin{figure}[h]
\begin{lstlisting}[language=nickel]
let f = fun x y => x in
let g = f | (Num -> Num -> Num)
            @& (Num -> Bool -> Num) in
let h = g 1 in
(h 1, h true)
\end{lstlisting}
\caption{Intersection contracts don't distribute}
\label{fig:intersection-distrbution}
\end{figure}

This discrepancy is inherent to the chosen semantics for intersections.  When
applying a function with an intersection contract, only one fixed component of
the intersection can be selected for the function.
one components of the intersection. Not that it must necessarily known right at the
application, but the result can't be used
functions, the choice of the component is fixed once for all when the function
is applied. In Figure~\ref{fig:intersection-distrbution}, the contract for
\nickel{h} can be either \nickel{Num -> Num} or \nickel{Bool -> Num}, but can't
be both. This is a manifestation of the following more general fact: the
distributivity law given in Figure~\ref{fig:distributivity-law} is not valid in
the semantics, while it is both intuitive and sound.\unsure{What means sound, in
this context?}

%
% It is worth noting, that intersection types do have this property,
% since subtyping would allow to correctly type check a term like
% \nickel{g 1} as both \nickel{Num -> Num} as well as
% \nickel{Bool -> Num}, and then deducing that \nickel{g 1}
% can be given the type \nickel{Num -> Num @& Bool -> Num}.
%

>>>>

A natural - but naive - implementation of intersection contracts could be the
following: to apply a contract \nickel{A @& B}, apply both contracts \nickel{A}
and \nickel{B} sequentially, resulting in the naive decomposition rule of
Figure~\ref{fig:naive-impl}.

\begin{figure}[h]
\begin{lstlisting}[language=nickel,frame=none,numbers=none,title={Naive
decomposition}]
M | A @& B (*@ $\simeq$ @*) (M | A) | B
\end{lstlisting}
\begin{lstlisting}[language=nickel,frame=none,numbers=none,title={Exchange law}]
(A -> B) @& (C -> D) (*@ $\simeq$ @*) (A @& C) -> (B @& D)
\end{lstlisting}
\begin{lstlisting}[language=nickel,title={Overloaded identity}]
let g | Num -> Num @& Str -> Str
      = fun x => x in
g 1
\end{lstlisting}
\caption{Naive implementation of intersection}
\label{fig:naive-impl}
\end{figure}

This intuition works for simple contracts: checking that \nickel{x | Natural @&
Odd} amounts to check that \nickel{x | Natural} and \nickel{x | Odd}.
Unfortunately, this doesn't scale to higher-order contracts.  The overloaded
identity example of Figure~\ref{fig:naive-impl} illustrates the use of an
intersection to model a simple overloading of the identity function. If we were
to apply the naive decomposition, the argument \nickel{1} would fail the
\nickel{Str -> Str} contract and abort the execution. Perhaps the exchange rule
given in Figure~\ref{fig:naive-impl}, which is a direct consequence of the naive
decomposition, illustrates the issue better. It is clear that this exchange law
isn't the right semantics for overloading. With this law, the contract for
overloaded identity of Figure~\ref{fig:naive-impl} would always fail because
no argument can satisfy \nickel{Num @& Str}.

@@

But which semantics? We deemed the exchange law of Figure~\ref{fig:naive-impl}
as "clearly wrong" with respect to overloading. There is nothing wrong per se :
we could chose a semantics for intersections that validate this rule.  However,
the resulting system wouldn't capture overloading, and is arguably not very
interesting nor useful. In~\cite{KeilThiemannUnionIntersection}, the authors
argue for basing the semantics of contracts on union and intersection type
systems, which have a simple and natural definition from which the other
properties follow. This approach makes sense, and has been continued in future
works, and in particular in~\cite{RootCauseOfBlame}.

But even this semantics is no panacea. Figure~\ref{fig:intersection-distribution}
shows an example that intuitively looks like it should succeeed, returning
\nickel{(1, 1)}. However, under the semantics used both
in~\cite{KeilThiemannUnionIntersection} and~\cite{RootCauseOfBlame}, this program
fails with a contract violation.

\begin{figure}[h]
\begin{lstlisting}[language=nickel]
let f = fun x y => x in
let g = f | (Num -> Num -> Num)
            @& (Num -> Bool -> Num) in
let h = g 1 in
(h 1, h true)
\end{lstlisting}
\caption{Intersection contracts don't distribute}
\label{fig:intersection-distribution}
\end{figure}

This discrepancy is inherent to the chosen semantics .  When applying a function
with an intersection contract, only one component of the intersection must be
selected for the function. The actual choice of component may be delayed to
later, but the point is that it must be the same across all usages. In
Figure~\ref{fig:intersection-distribution}, the contract for \nickel{h} can in
consequence be either \nickel{Num -> Num} or \nickel{Bool -> Num}, but not both.
This is a manifestation of the following more general defect: the distributivity
law given in Figure~\ref{fig:distributivity-law} is not valid in the semantics
given in~\cite{KeilThiemannUnionIntersection}, while it is both sound to add and
intuitive.

Curiously, union and intersection type systems, on which
\cite{KeilThiemannUnionIntersection} are basing their semantics on, don't
exhibit this property. Subtyping allows to correctly type check a term like
\nickel{g 1} as both \nickel{Num -> Num} as well as
\nickel{Bool -> Num}, and then deducing that \nickel{g 1}
can be given the type \nickel{Num -> Num @& Bool -> Num}.
\unsure{(Yann) I just reworked the wording, but I'm not totally sold on the
argument}

@@
